Using functor categories to generate intermediate code with Agda

### 2025-05-19 temp.tex
Minitoc is back! Hooray!

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & 9574 \footnotemark \\
        \textbf{Code line count:} & 1727 \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

    I defined a compiler in Agda for the source language and target language as described in Reynolds' work~\autocite{Reynolds}. The compiler was guaranteed to generate expected code by tests in a separate Agda file. An additional customised library was defined to support the implementation of the compiler. More importantly, I spotted a mistake in Reynolds' definition of the denotational semantics of the source language and provided a corrected version. All success criteria were met, and the project was completed successfully.

@@

    None.

    \begingroup
        \footnotetext[1]{This word-count was computed by \texttt{texcount -sum -1 content.tex}.}
        \footnotetext[2]{This code line count was computed by \texttt{find . -name "*.agda" | xargs wc -l}.}
    \endgroup
\end{titlepage}

@@

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory, which will be discussed in~\secref{subsec: curry-howard-lambek}.

@@

        \subsection{Category}
        \begin{definition}[Category] \label{def: category}
            A \emph{category} $\mathcal{C}$ is specified by
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \xrightarrow{\mathcal{C}} Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit)}
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity)}
                    For all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

        \begin{example}[$\mathcal{Set}$]
            An example of category is the category of sets, denoted as $\mathcal{Set}$, specified by the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{Set}$ are small\footnote{Here the smallness condition avoids foundational issues analogous to Russell's paradox. While we omit the formal definition of size due to the space limit here, all constructions in this work preserve smallness.} sets;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{Set})$, the morphisms from $X \to Y$ in $\mathcal{Set}$ are the functions $X \to Y$;
                \item
                    The identity morphism on $X$ in $\mathcal{Set}$ is the identity function on $X$;
                \item
                    The composition of morphisms in $\mathcal{Set}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{Set}$, since the composition of functions is associative and the identity function compose with any function is the function itself.
        \end{example}

        \begin{definition}[Preorder]
            A \emph{preorder} $\underline{P} = (P, \sqsubseteq)$ is a set $P$ with a binary relation $\sqsubseteq$ that is
            \begin{itemize}
                \item reflexive: $\forall x \in P, x \sqsubseteq x$;
                \item transitive: $\forall x, y, z \in P, x \sqsubseteq y \land y \sqsubseteq z \implies x \sqsubseteq z$.
            \end{itemize}
        \end{definition}

        \begin{example}[Category determined by preorder] \label{ex: category_preorder}
            Another example of category is a category $\mathcal{C}_{\underline{P}}$ determined by any preorder $\underline{P}$, which is specified by the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{C}_{\underline{P}}$ are the elements of $P$;
                \item
                    $\mathcal{C}_{\underline{P}}(x,y) = \begin{cases}
                        \{(x,y)\} & \text{if $x \sqsubseteq y$} \\
                        \emptyset & \text{otherwise}
                    \end{cases}$
                \item
                    The identity morphism on $X$ in $\mathcal{C}_{\underline{P}}$ is the identity function on $X$;
                \item
                    The composition of morphisms in $\mathcal{C}_{\underline{P}}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{C}_{\underline{P}}$.

            $\mathcal{C}_{\underline{P}}$ is used later for stack descriptors in the denotational semantics of the source language.
        \end{example}

        The idea of opposite category is that if we have a category $\mathcal{C}$, we can reverse the direction of all morphisms in $\mathcal{C}$ to obtain a new category $\mathcal{C}^{\textbf{op}}$.

        \begin{definition}[Opposite category] \label{def: opposite}
            Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                \item 
                    For each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                \item 
                    The identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                \item 
                    The composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
            \end{itemize}
        \end{definition}

            The notation of commutative diagram is widely used in category theory as a convenient visual representation of the relationships between objects and morphisms in a category.
            \subsubsection{Commutative diagrams}
            A \emph{diagram} in a category $\mathcal{C}$ is a directed graph whose vertices are $\mathcal{C}$-objects and whose edges are $\mathcal{C}$-morphisms.

            A diagram is \emph{commutative} (or \emph{commutes}) if any two finite paths in the graph between any two vertices $X$ and $Y$ in the diagram determine the equal morphism $f \in \mathcal{C}{(X,Y)}$ under the composition of morphisms.

@@

        \subsection{Isomorphism}
        \begin{definition}[Isomorphism]
            Given a category $\mathcal{C}$, a $\mathcal{C}$-morphism $f : X \xrightarrow{\mathcal{C}} Y$ is called an \emph{isomorphism} if there exists a morphism $g : Y \xrightarrow{\mathcal{C}} X$ such that the following diagram commutes:
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{isomorphism.pdf}}}
                \label{cd: isomorphism}
            \end{equation}
            In other words, $f$ is an isomorphism if there exists a morphism $g$ such that $g \circ f = \textbf{id}_X$ and $f \circ g = \textbf{id}_Y$.

            The morphism $g$ is uniquely determined by $f$ and is called the \emph{inverse} of $f$, denoted as $f^{-1}$.

            Given two objects $X$ and $Y$ in a category $\mathcal{C}$, if there exists an isomorphism from $X$ to $Y$, we say that $X$ and $Y$ are \emph{isomorphic} in $\mathcal{C}$ and write $X \cong Y$.

@@

        \subsection{Cartesian closed category}
            \begin{definition}[Terminal object]
                Given a category $\mathcal{C}$, an object $T \in \textbf{obj}(\mathcal{C})$ is called a \emph{terminal object} if for all $X \in \textbf{obj}(\mathcal{C})$, there exists a unique \textbf{C}-morphism $f : X \to T$.
            \end{definition}
            Terminal objects are unique up to isomorphism. In other words, we have the following properties:
            \begin{itemize}
                \item 
                    If $T$ and $T'$ are both terminal objects in $\mathcal{C}$, then there exists a unique isomorphism $f : T \to T'$.
                \item
                    If $T$ is a terminal object in $\mathcal{C}$ and $T \cong T'$, then $T'$ is also a terminal object in $\mathcal{C}$.
            \end{itemize}
            \begin{example}[Terminal object in $\mathcal{Set}$]
                $\mathcal{Set}$ has a terminal object $\{\ast\}$, which is an arbitrary singleton set containing a single element $\ast$.

                For any set $X$, there exists a unique function $f : X \to \{\ast\}$ that maps every element of $X$ to the single element $\ast$ in $\{\ast\}$.

                There is a unique isomorphism $f : \{\ast\} \to \{\cdot\}$ for any two singleton sets $\{\ast\}$ and $\{\cdot\}$, which is $f(\ast) = \cdot$.
            \end{example}

            \begin{definition}[Binary product]
                Given a category $\mathcal{C}$, the \emph{binary product} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \times Y$;
                    \item two $\mathcal{C}$-morphisms $\pi_1 : X \times Y \to X$ and $\pi_2 : X \times Y \to Y$ called the \emph{projections} of $X \times Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \to X$ and $g : Z \to Y$, there exists a unique morphism $u : Z \to X \times Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{binary_product.pdf}}}
                    \label{cd: binary_product}
                \end{equation}
                The unique morphism $u$ is written as $\ang{ f, g } : Z \to X \times Y$ where $f = \pi_1 \circ u$ and $g = \pi_2 \circ u$.
            \end{definition}
            It can be shown that the binary product is unique up to (unique) isomorphism.

            \begin{example}[Binary product in $\mathcal{Set}$]
                The binary product of two sets $X$ and $Y$ in $\mathcal{Set}$ is the cartesian product $X \times Y = \{(x,y) \mid x \in X \land y \in Y\}$, where $(x,y)$ are ordered pairs.

                We have the following projections:
                \begin{itemize}
                    \item 
                        $\pi_1 : X \times Y \to X$ is defined as $\pi_1(x,y) = x$ for all $(x,y) \in X \times Y$;
                    \item
                        $\pi_2 : X \times Y \to Y$ is defined as $\pi_2(x,y) = y$ for all $(x,y) \in X \times Y$.
                \end{itemize}

                For any set $Z$, for any functions $f : Z \to X$ and $g : Z \to Y$, the unique morphism $u : Z \to X \times Y$ is defined as:
                \[u(z) = (f(z), g(z)) \text{ for all $z \in Z$.}\]
            \end{example}

            \begin{definition}[Exponential]
                Given a category $\mathcal{C}$ with binary products, the \emph{exponential} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \Rightarrow Y$;
                    \item a $\mathcal{C}$-morphism $\textbf{app} : (X \Rightarrow Y) \times X \to Y$ called the \emph{application} of $X \Rightarrow Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \times X \to Y$, there exists a unique morphism $u : Z \to X \Rightarrow Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{exponential.pdf}}}
                    \label{cd: exponential}
                \end{equation}
                We write $\textbf{cur} f$ for the unique morphism $u$ such that $f = \textbf{app} \circ (\textbf{cur} f \times \textbf{id}_X)$, 
                where $\textbf{cur} f$ is called the \emph{currying} of $f$.
            \end{definition}
            It can be shown that the exponential is unique up to (unique) isomorphism.

            \begin{example}[Exponential in $\mathcal{Set}$]
                The exponential of two sets $X$ and $Y$ in $\mathcal{Set}$ is the set of all functions from $X$ to $Y$.

                Function application gives the morphism $\textbf{app} : (X \Rightarrow Y) \times X \to Y$ as $\textbf{app}(f,x) = f(x)$ for all $f \in X \Rightarrow Y$ and $x \in X$.

                The currying operation transform a function $f : Z \times X \to Y$ into a function $\textbf{cur} f : Z \to (X \Rightarrow Y)$, which is defined as $\textbf{cur} f(z) = \lambda x.f(z,x)$ for all $z \in Z$ and $x \in X$.
            \end{example}

        \begin{definition}[Cartesian closed category]
            A category $\mathcal{C}$ is called a \emph{cartesian closed category} (CCC) if it has a terminal object, binary products and exponentials of any two objects.
        \end{definition}

        \subsection{Functor}
        \begin{definition}[Functor] \label{def: functor}
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by
            \begin{itemize}
                \item 
                    a function referred to as object mapping
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function referred to as functorial mapping
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have: $ F(\textbf{id}_X) = \textbf{id}_{F(X)}$
                \item
                    For all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have: $F(g \circ f) = F(g) \circ F(f)$
            \end{itemize}
        \end{definition}

@@

            Given two categories $\mathcal{C}$ and $\mathcal{D}$, and two functors $F,G: \mathcal{C} \to \mathcal{D}$, a \emph{natural transformation} $\theta : F \to G$ is a family of morphisms $\theta_X \in \mathcal{D}{(F(X),G(X))}$ for each $X \in \textbf{obj}(\mathcal{C})$ such that for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, the following diagram 
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{natural_transformation.pdf}}}
                \label{cd: natural_transformation}
            \end{equation}

            commutes in $\mathcal{D}$, i.e. $G(f) \circ \theta_X = \theta_Y \circ F(f)$

@@

        \subsection{Functor category}
        \begin{definition}[Functor category]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, the \emph{functor category} $\mathcal{D}^{\mathcal{C}}$ is the category satisfying the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{D}^{\mathcal{C}}$ are all functors $\mathcal{C} \to \mathcal{D}$;
                \item 
                    Given two functors $F,G: \mathcal{C} \to \mathcal{D}$, the morphisms from $F$ to $G$ in $\mathcal{D}^{\mathcal{C}}$ are all natural transformations $\theta : F \to G$;
                \item
                    Composition and identity morphisms in $\mathcal{D}^{\mathcal{C}}$ are defined as follows:
                    \begin{itemize}
                        \item 
                            The identity morphism $\textbf{id}_F$ on $F$ is defined as $\theta_X = \textbf{id}_{F(X)}$ for all $X \in \textbf{obj}(\mathcal{C})$;
                        \item
                            The composition of two natural transformations $\theta : F \to G$ and $\phi : G \to H$ is defined as $(\phi \circ \theta)_X = \phi_X \circ \theta_X$ for all $X \in \textbf{obj}(\mathcal{C})$.
                    \end{itemize}
            \end{itemize}
        \end{definition}

>>>>

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & [wordcount] \footnotemark \\
        \textbf{Code line count:} & [linecount] \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

    I have met all my proposed success criteria and have implemented three extensions. A formalisation of source language and target language in Agda was provided. The compiler was implemented as an Agda file that takes a source program and produces a target program. The compiler was guaranteed to generate expected code by tests in a separate Agda file. An additional customised library was defined to support the implementation of the compiler. More importantly, I spotted a mistake in Reynolds' definition of the denotational semantics of the source language and provided a corrected version.

@@

    The implementation depends on the theory provided by Reynolds~\autocite{Reynolds}. Some details in the theory were missing, and a type definition in the theory was incorrect.

    \begingroup
        \footnotetext[1]{This word-count was computed via \texttt{texcount}.}
        \footnotetext[2]{This code line count was computed by \texttt{find . -name "*.agda" | xargs wc -l}.}
    \endgroup
\end{titlepage}

@@

        Notation in category theory is similar to that in set theory and type theory. For example, \text{A, B, \dots} are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ \text{A \to B}). There is a deeper connection between category theory and type theory, which will be discussed in~\secref{subsec: curry-howard-lambek}.

@@

        \subsection{Category}
        \begin{definition}[Category] \label{def: category}
            A \emph{category} \text{\mathcal{C}} is specified by
            \begin{itemize}
                \item 
                    a collection of objects \text{\textbf{obj}(\mathcal{C})}, whose elements are called \text{\mathcal{C}}-objects;
                \item 
                    for each \text{X, Y \in \textbf{obj}(\mathcal{C})}, a collection of morphisms \text{\mathcal{C}{(X,Y)}}, whose elements are called \text{\mathcal{C}}-morphisms from \text{X} to \text{Y} (e.g.\ \text{f : X \xrightarrow{\mathcal{C}} Y});
                \item 
                    for each \text{X \in \textbf{obj}(\mathcal{C})}, an element \text{\textbf{id}_X \in \mathcal{C}{(X,X)}} called the identity morphism on \text{X};
                \item 
                    for each \text{X, Y, Z \in \textbf{obj}(\mathcal{C})}, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit)}
                    For all \text{X, Y \in \textbf{obj}(\mathcal{C})} and \text{f \in \mathcal{C}{(X,Y)}}, we have
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity)}
                    For all \text{X, Y, Z, W \in \textbf{obj}(\mathcal{C})} and \text{f \in \mathcal{C}{(X,Y)}}, \text{g \in \mathcal{C}{(Y,Z)}}, \text{h \in \mathcal{C}{(Z,W)}}, we have
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

        \begin{example}[\text{\mathcal{Set}}]
            An example of category is the category of sets, denoted as \text{\mathcal{Set}}, specified by the following:
            \begin{itemize}
                \item 
                    The objects of \text{\mathcal{Set}} are small\footnote{Here the smallness condition avoids foundational issues analogous to Russell's paradox. While we omit the formal definition of size due to the space limit here, all constructions in this work preserve smallness.} sets;
                \item
                    For each \text{X, Y \in \textbf{obj}(\mathcal{Set})}, the morphisms from \text{X \to Y} in \text{\mathcal{Set}} are the functions \text{X \to Y};
                \item
                    The identity morphism on \text{X} in \text{\mathcal{Set}} is the identity function on \text{X};
                \item
                    The composition of morphisms in \text{\mathcal{Set}} is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in \text{\mathcal{Set}}, since the composition of functions is associative and the identity function compose with any function is the function itself.
        \end{example}

        \begin{definition}[Preorder]
            A \emph{preorder} \text{\underline{P} = (P, \sqsubseteq)} is a set \text{P} with a binary relation \text{\sqsubseteq} that is
            \begin{itemize}
                \item reflexive: \text{\forall x \in P, x \sqsubseteq x};
                \item transitive: \text{\forall x, y, z \in P, x \sqsubseteq y \land y \sqsubseteq z \implies x \sqsubseteq z}.
            \end{itemize}
        \end{definition}

        \begin{example}[Category determined by preorder] \label{ex: category_preorder}
            Another example of category is a category \text{\mathcal{C}_{\underline{P}}} determined by any preorder \text{\underline{P}}, which is specified by the following:
            \begin{itemize}
                \item 
                    The objects of \text{\mathcal{C}_{\underline{P}}} are the elements of \text{P};
                \item
                    $\mathcal{C}_{\underline{P}}(x,y) = \begin{cases}
                        \{(x,y)\} & \text{if \text{x \sqsubseteq y}} \\
                        \emptyset & \text{otherwise}
                    \end{cases}$
                \item
                    The identity morphism on \text{X} in \text{\mathcal{C}_{\underline{P}}} is the identity function on \text{X};
                \item
                    The composition of morphisms in \text{\mathcal{C}_{\underline{P}}} is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in \text{\mathcal{C}_{\underline{P}}}.

            \text{\mathcal{C}_{\underline{P}}} is used later for stack descriptors in the denotational semantics of the source language.
        \end{example}

        The idea of opposite category is that if we have a category \text{\mathcal{C}}, we can reverse the direction of all morphisms in \text{\mathcal{C}} to obtain a new category \text{\mathcal{C}^{\textbf{op}}}.

        \begin{definition}[Opposite category] \label{def: opposite}
            Given a category \text{\mathcal{C}}, its \emph{opposite category} \text{\mathcal{C}^{\textbf{op}}} is specified by the following:
            \begin{itemize}
                \item 
                    The objects of \text{\mathcal{C}^{\textbf{op}}} are the same as those of \text{\mathcal{C}};
                \item 
                    For each \text{X, Y \in \textbf{obj}(\mathcal{C})}, the morphisms from \text{X} to \text{Y} in \text{\mathcal{C}^{\textbf{op}}} are the morphisms from \text{Y} to \text{X} in \text{\mathcal{C}};
                \item 
                    The identity morphism on \text{X} in \text{\mathcal{C}^{\textbf{op}}} is the identity morphism on \text{X} in \text{\mathcal{C}};
                \item 
                    The composition of morphisms in \text{\mathcal{C}^{\textbf{op}}} is defined as the composition of morphisms in \text{\mathcal{C}}.
            \end{itemize}
        \end{definition}

            The notation of commutative diagram is widely used in category theory as a convenient visual representation of the relationships between objects and morphisms in a category.
            \subsubsection{Commutative diagrams}
            A \emph{diagram} in a category \text{\mathcal{C}} is a directed graph whose vertices are \text{\mathcal{C}}-objects and whose edges are \text{\mathcal{C}}-morphisms.

            A diagram is \emph{commutative} (or \emph{commutes}) if any two finite paths in the graph between any two vertices \text{X} and \text{Y} in the diagram determine the equal morphism \text{f \in \mathcal{C}{(X,Y)}} under the composition of morphisms.

@@

        \subsection{Isomorphism}
        \begin{definition}[Isomorphism]
            Given a category \text{\mathcal{C}}, a \text{\mathcal{C}}-morphism \text{f : X \xrightarrow{\mathcal{C}} Y} is called an \emph{isomorphism} if there exists a morphism \text{g : Y \xrightarrow{\mathcal{C}} X} such that the following diagram commutes:
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{isomorphism.pdf}}}
                \label{cd: isomorphism}
            \end{equation}
            In other words, \text{f} is an isomorphism if there exists a morphism \text{g} such that \text{g \circ f = \textbf{id}_X} and \text{f \circ g = \textbf{id}_Y}.

            The morphism \text{g} is uniquely determined by \text{f} and is called the \emph{inverse} of \text{f}, denoted as \text{f^{-1}}.

            Given two objects \text{X} and \text{Y} in a category \text{\mathcal{C}}, if there exists an isomorphism from \text{X} to \text{Y}, we say that \text{X} and \text{Y} are \emph{isomorphic} in \text{\mathcal{C}} and write \text{X \cong Y}.

@@

        \subsection{Cartesian closed category}
            \begin{definition}[Terminal object]
                Given a category \text{\mathcal{C}}, an object \text{T \in \textbf{obj}(\mathcal{C})} is called a \emph{terminal object} if for all \text{X \in \textbf{obj}(\mathcal{C})}, there exists a unique \textbf{C}-morphism \text{f : X \to T}.
            \end{definition}
            Terminal objects are unique up to isomorphism. In other words, we have the following properties:
            \begin{itemize}
                \item 
                    If \text{T} and \text{T'} are both terminal objects in \text{\mathcal{C}}, then there exists a unique isomorphism \text{f : T \to T'}.
                \item
                    If \text{T} is a terminal object in \text{\mathcal{C}} and \text{T \cong T'}, then \text{T'} is also a terminal object in \text{\mathcal{C}}.
            \end{itemize}
            \begin{example}[Terminal object in \text{\mathcal{Set}}]
                \text{\mathcal{Set}} has a terminal object \text{\{\ast\}}, which is an arbitrary singleton set containing a single element \text{\ast}.

                For any set \text{X}, there exists a unique function \text{f : X \to \{\ast\}} that maps every element of \text{X} to the single element \text{\ast} in \text{\{\ast\}}.

                There is a unique isomorphism \text{f : \{\ast\} \to \{\cdot\}} for any two singleton sets \text{\{\ast\}} and \text{\{\cdot\}}, which is \text{f(\ast) = \cdot}.
            \end{example}

            \begin{definition}[Binary product]
                Given a category \text{\mathcal{C}}, the \emph{binary product} of two objects \text{X} and \text{Y} in \text{\mathcal{C}} is specified by
                \begin{itemize}
                    \item a \text{\mathcal{C}}-object \text{X \times Y};
                    \item two \text{\mathcal{C}}-morphisms \text{\pi_1 : X \times Y \to X} and \text{\pi_2 : X \times Y \to Y} called the \emph{projections} of \text{X \times Y};
                \end{itemize}
                such that for all \text{Z \in \textbf{obj}(\mathcal{C})} and morphisms \text{f : Z \to X} and \text{g : Z \to Y}, there exists a unique morphism \text{u : Z \to X \times Y} such that the following diagram commutes in \text{\mathcal{C}}:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{binary_product.pdf}}}
                    \label{cd: binary_product}
                \end{equation}
                The unique morphism \text{u} is written as \text{\ang{ f, g } : Z \to X \times Y} where \text{f = \pi_1 \circ u} and \text{g = \pi_2 \circ u}.
            \end{definition}
            It can be shown that the binary product is unique up to (unique) isomorphism.

            \begin{example}[Binary product in \text{\mathcal{Set}}]
                The binary product of two sets \text{X} and \text{Y} in \text{\mathcal{Set}} is the cartesian product \text{X \times Y = \{(x,y) \mid x \in X \land y \in Y\}}, where \text{(x,y)} are ordered pairs.

                We have the following projections:
                \begin{itemize}
                    \item 
                        \text{\pi_1 : X \times Y \to X} is defined as \text{\pi_1(x,y) = x} for all \text{(x,y) \in X \times Y};
                    \item
                        \text{\pi_2 : X \times Y \to Y} is defined as \text{\pi_2(x,y) = y} for all \text{(x,y) \in X \times Y}.
                \end{itemize}

                For any set \text{Z}, for any functions \text{f : Z \to X} and \text{g : Z \to Y}, the unique morphism \text{u : Z \to X \times Y} is defined as:
                \[u(z) = (f(z), g(z)) \text{ for all \text{z \in Z}.}\]
            \end{example}

            \begin{definition}[Exponential]
                Given a category \text{\mathcal{C}} with binary products, the \emph{exponential} of two objects \text{X} and \text{Y} in \text{\mathcal{C}} is specified by
                \begin{itemize}
                    \item a \text{\mathcal{C}}-object \text{X \Rightarrow Y};
                    \item a \text{\mathcal{C}}-morphism \text{\textbf{app} : (X \Rightarrow Y) \times X \to Y} called the \emph{application} of \text{X \Rightarrow Y};
                \end{itemize}
                such that for all \text{Z \in \textbf{obj}(\mathcal{C})} and morphisms \text{f : Z \times X \to Y}, there exists a unique morphism \text{u : Z \to X \Rightarrow Y} such that the following diagram commutes in \text{\mathcal{C}}:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{exponential.pdf}}}
                    \label{cd: exponential}
                \end{equation}
                We write \text{\textbf{cur} f} for the unique morphism \text{u} such that \text{f = \textbf{app} \circ (\textbf{cur} f \times \textbf{id}_X)}, 
                where \text{\textbf{cur} f} is called the \emph{currying} of \text{f}.
            \end{definition}
            It can be shown that the exponential is unique up to (unique) isomorphism.

            \begin{example}[Exponential in \text{\mathcal{Set}}]
                The exponential of two sets \text{X} and \text{Y} in \text{\mathcal{Set}} is the set of all functions from \text{X} to \text{Y}.

                Function application gives the morphism \text{\textbf{app} : (X \Rightarrow Y) \times X \to Y} as \text{\textbf{app}(f,x) = f(x)} for all \text{f \in X \Rightarrow Y} and \text{x \in X}.

                The currying operation transform a function \text{f : Z \times X \to Y} into a function \text{\textbf{cur} f : Z \to (X \Rightarrow Y)}, which is defined as \text{\textbf{cur} f(z) = \lambda x.f(z,x)} for all \text{z \in Z} and \text{x \in X}.
            \end{example}

        \begin{definition}[Cartesian closed category]
            A category \text{\mathcal{C}} is called a \emph{cartesian closed category} (CCC) if it has a terminal object, binary products and exponentials of any two objects.
        \end{definition}

        \subsection{Functor}
        \begin{definition}[Functor] \label{def: functor}
            Given two categories \text{\mathcal{C}} and \text{\mathcal{D}}, a \emph{functor} \text{F: \mathcal{C} \to \mathcal{D}} is specified by
            \begin{itemize}
                \item 
                    a function referred to as object mapping
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each \text{X, Y \in \textbf{obj}(\mathcal{C})}, a function referred to as functorial mapping
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    For all \text{X, Y \in \textbf{obj}(\mathcal{C})} and \text{f \in \mathcal{C}{(X,Y)}}, we have: \text{ F(\textbf{id}_X) = \textbf{id}_{F(X)}}
                \item
                    For all \text{X, Y, Z \in \textbf{obj}(\mathcal{C})} and \text{f \in \mathcal{C}{(X,Y)}}, \text{g \in \mathcal{C}{(Y,Z)}}, we have: \text{F(g \circ f) = F(g) \circ F(f)}
            \end{itemize}
        \end{definition}

@@

            Given two categories \text{\mathcal{C}} and \text{\mathcal{D}}, and two functors \text{F,G: \mathcal{C} \to \mathcal{D}}, a \emph{natural transformation} \text{\theta : F \to G} is a family of morphisms \text{\theta_X \in \mathcal{D}{(F(X),G(X))}} for each \text{X \in \textbf{obj}(\mathcal{C})} such that for all \text{X, Y \in \textbf{obj}(\mathcal{C})} and \text{f \in \mathcal{C}{(X,Y)}}, the following diagram 
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{natural_transformation.pdf}}}
                \label{cd: natural_transformation}
            \end{equation}

            commutes in \text{\mathcal{D}}, i.e. \text{G(f) \circ \theta_X = \theta_Y \circ F(f)}

@@

        \subsection{Functor category}
        \begin{definition}[Functor category]
            Given two categories \text{\mathcal{C}} and \text{\mathcal{D}}, the \emph{functor category} \text{\mathcal{D}^{\mathcal{C}}} is the category satisfying the following:
            \begin{itemize}
                \item 
                    The objects of \text{\mathcal{D}^{\mathcal{C}}} are all functors \text{\mathcal{C} \to \mathcal{D}};
                \item 
                    Given two functors \text{F,G: \mathcal{C} \to \mathcal{D}}, the morphisms from \text{F} to \text{G} in \text{\mathcal{D}^{\mathcal{C}}} are all natural transformations \text{\theta : F \to G};
                \item
                    Composition and identity morphisms in \text{\mathcal{D}^{\mathcal{C}}} are defined as follows:
                    \begin{itemize}
                        \item 
                            The identity morphism \text{\textbf{id}_F} on \text{F} is defined as \text{\theta_X = \textbf{id}_{F(X)}} for all \text{X \in \textbf{obj}(\mathcal{C})};
                        \item
                            The composition of two natural transformations \text{\theta : F \to G} and \text{\phi : G \to H} is defined as \text{(\phi \circ \theta)_X = \phi_X \circ \theta_X} for all \text{X \in \textbf{obj}(\mathcal{C})}.
                    \end{itemize}
            \end{itemize}
        \end{definition}
